---
title: "A6_ATReyn"
output: html_notebook
---

### Setup

Importing libraries:
```{r, echo = F}

library(BiocManager)
library(genbankr)
library(Biostrings)
library(ggplot2)
library(ggtree)
library(annotate)
library(muscle)
library(reshape2)
library(rentrez)
library(ape)
library(ggpubr)

```

Importing and cleaning data:
```{r}

# Import data
dat <- "> Human isolate, unknown sequence
ATGTCTGATAATGGACCCCAAAATCAGCGAAATGCACCCCGCATTACGTTTGGTGGACCCTCAGATTCAA
CTGGCAGTAACCAGAATGGAGAACGCAGTGGGGCGCGATCAAAACAACGTCGGCCCCAAGGTTTACCCAA
TAATACTGCGTCTTGGTTCACCGCTCTCACTCAACATGGCAAGGAAGACCTTAAATTCCCTCGAGGACAA
GGCGTTCCAATTAACACCAATAGCAGTCCAGATGACCAAATTGGCTACTACCGAAGAGCTACCAGACGAA
TTCGTGGTGGTGACGGTAAAATGAAAGATCTCAGTCCAAGATGGTATTTCTACTACCTAGGAACTGGGCC
AGAAGCTGGACTTCCCTATGGTGCTAACAAAGACGGCATCATATGGGTTGCAACTGAGGGAGCCTTGAAT
ACACCAAAAGATCACATTGGCACCCGCAATCCTGCTAACAATGCTGCAATCGTGCTACAACTTCCTCAAG
GAACAACATTGCCAAAAGGCTTCTACGCAGAAGGGAGCAGAGGCGGCAGTCAAGCCTCTTCTCGTTCCTC
ATCACGTAGTCGCAACAGTTCAAGAAATTCAACTCCAGGCAGCAGTAGGGGAACTTCTCCTGCTAGAATG
GCTGGCAATGGCGGTGATGCTGCTCTTGCTTTGCTGCTGCTTGACAGATTGAACCAGCTTGAGAGCAAAA
TGTCTGGTAAAGGCCAACAACAACAAGGCCAAACTGTCACTAAGAAATCTGCTGCTGAGGCTTCTAAGAA
GCCTCGGCAAAAACGTACTGCCACTAAAGCATACAATGTAACACAAGCTTTCGGCAGACGTGGTCCAGAA
CAAACCCAAGGAAATTTTGGGGACCAGGAACTAATCAGACAAGGAACTGATTACAAACATTGGCCGCAAA
TTGCACAATTTGCCCCCAGCGCTTCAGCGTTCTTCGGAATGTCGCGCATTGGCATGGAAGTCACACCTTC
GGGAACGTGGTTGACCTACACAGGTGCCATCAAATTGGATGACAAAGATCCAAATTTCAAAGATCAAGTC
ATTTTGCTGAATAAGCATATTGACGCATACAAAACATTCCCACCAACAGAGCCTAAAAAGGACAAAAAGA
AGAAGGCTGATGAAACTCAAGCCTTACCGCAGAGACAGAAGAAACAGCAAACTGTGACTCTTCTTCCTGC
TGCAGATTTGGATGATTTCTCCAAACAATTGCAACAATCCATGAGCAGTGCTGACTCAACTCAGGCCTAA"

# Tidy data

seq <- gsub("^>.*sequence\\n([ATCG].*)", "\\1", dat)
seq <- gsub("\\n", "", seq)

```
\n

### Pairwise Alignments

Finding pairwise alignments between our query search (seq) and hits from NCBI database:
```{r}

# BLAST search
seqBLAST <- blastSequences(paste(seq), as = 'data.frame',
                           hitListSize = 40, timeout = 600)

```
\n

### Multiple Alignments

Examining hits from blast search:
```{r}

# Load accession numbers from hits into dataframe object
seqHitsDF <- data.frame(ID = seqBLAST$Hit_accession, Seq = seqBLAST$Hsp_hseq,
                     stringsAsFactors = FALSE)

# Preview hit lengths for each hit
seqBLAST$Hit_len

# Preview species matches
seqHitSp <- read.GenBank(seqBLAST$Hit_accession)
attr(seqHitSp, "species")

```
It appears that all the matches for the human isolate from the patient biofluid sample belong to SARS-COV-2.
\n

### Aligning Sequences

Prepping data for alignment
```{r}

# Convert sequences into DNAString object
seqDNAString <- seqHitsDF$Seq %>% 
  as.character %>% # Convert to strings
  lapply(.,paste0, collapse = "") %>% # Convert into single string
  unlist %>% # Remove list so it becomes a single vector
  DNAStringSet  # Convert vector to DNAString

# Since each hit belongs to SARS-COV-2, we will create unique names for each sequence so they can be distinguished
names(seqDNAString) <- paste(1:nrow(seqHitsDF), seqHitsDF$ID, sep = "_")

```
\n

Create alignment file:
```{r}

# Run alignment with muscle from muscle package
seqAlign <- muscle::muscle(stringset = seqDNAString, quiet = T)

```
\n

### Inspecting Alignment

Examining sequences alignment:
```{r}

print(seqAlign)

```
There are not many large gaps in data, so we will not have to remove gaps.

\n

Rerunning alignment to sure that no large gaps are present:
```{r}

# Rerun alignment
seqSubAlign <- muscle::muscle(seqDNAString, quiet = T)

# Preview alignment
print(seqSubAlign)

```
Looks good!

\n

### Distance Matrix

A distance matrix can be used to determine the evolutionary relationships between sequences.

Setup:
```{r}

# Since we will be using ape package, need to convert to DNABin
seqAlignBin <- as.DNAbin(seqAlign)

# Creating distance matrix
seqDM <- dist.dna(seqAlignBin, model = "K80") # Evolutionary model designated
class(seqDM) # Verify conversion was successful (class = dist)

# Rearranging matrix to a "linear" matrix
seqDMMat <- as.matrix(seqDM)
dim(seqDMMat)

# Using melt to reshape data
seqMat <- melt(seqDMMat)
dim(seqMat)


```
\n

Visualizing the matrix:
```{r}

# Creating tile plot
fig1 <- ggplot(data = seqMat, aes(x = Var1, y = Var2, fill = value)) + 
  geom_tile() +
  scale_fill_gradientn(colours = c("white", "blue", "green", "red")) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))

# Print plot
print(fig1)

```
\n

### Creating Relationship Trees

Neighbour-joining method:
```{r}

# Setting up NJ method
seqTreeNJ <- nj(seqDM)
str(seqTreeNJ) # Checking that setup was successful

# Creating tree
timeplot <- ggtree(seqTreeNJ) +
  geom_tiplab()

print(timeplot)

# Previewing tree without length info to highlight relationships
notimeplot <- ggtree(seqTreeNJ, branch.length = 'none') +
  geom_tiplab() +
  xlim(NA, 50)

print(notimeplot)

# Joining plots
fig2 <- ggarrange(timeplot, notimeplot,
                  labels = c("A", "B"),
                  ncol = 2, nrow = 1)
fig2


```
\n

### Conclusions

The non-human DNA isolated from the patient's biofluid sample belongs to SARS-COV-2. A BLAST search and subsequent DNA alignment showed a 100% match between the sample and all 40 of its hits (Fig. 1). Additionally, plotting the evolutionary relationships between the sample and its hits produced a phylogeny in which each sample is very closely related, with evolutions occurring over very short timescales (Fig. 2). This evidences that the patient has been in contact with SARS-COV-2, although further testing may be required to determine if a coronavirus infection is responsible for their life-threatening illness. The DNA alignment and phylogeny both show that the SARS-COV-2 variation the patient was exposed to is not a new strain or mutation, as it has a 100% match with other SARS-COV-2 variations that have been sequenced. Although SARS-COV-2 infection is a serious concern, and may be contributing to the patient's decline in health, since it is not a novel mutation regular SARS-COV-2 procedures can be carried out when treating this patient. 

\n
```{r, echo = F}

print(fig1)

```
**Fig. 1.** A tile plot illustrating a matrix DNA alignment between the SARS-COV-2 DNA sample taken from the patient's biofliuds and its 40 matches found during a BLAST search. Colour denotes distances between alignments. No differences were found between any samples.

\n
\n

```{r, echo = F}

print(fig2)

```
**Fig. 2.** A phylogeny showing the evolutionary relationships between the SARS-COV-2 DNA sample taken from the patient's biofluids and its 40 matches found during a BLAST search. (**A**) All samples are very closely related and occurred over a short timescale. Branch length denotes time. (**B**) The same phylogeny with time relationships removed to highlight evolutionary relationships between samples. The basic, repetitive pattern indicates that the samples are very closely related and indistinguishable (this tree represents only one of many possible evolutionary trees).











































